Loop-Aware Optimizations in PyPy's Tracing JIT

### 2012-08-07 talk/dls2012/paper.tex
typos

\begin{abstract}
One of the nice properties of a tracing JIT is that many of its optimization
are simple requiring one forward pass only. This is not true for loop-invariant code
motion which is a very important optimization for code with tight kernels.
Especially for dynamic languages that typically performs quite a lot of loop invariant
type checking, boxed value unwrapping and virtual method lookups.
In this paper we present a scheme for making simple optimizations loop-aware by
using a simple pre-processing step on the trace and not changing the
optimizations themselves. The scheme can give performance improvements of a
factor over two for PyPy's Python JIT executing simple numerical kernels
bringing the performance close to that of compiled C code.
\end{abstract}

@@

A dynamic language typically needs to do quite a lot of type
checking, wrapping/unwrapping of boxed values, and virtual method dispatching. 
For tight computationally intensive loops a
significant amount of the execution time might be spend on such tasks
instead of the actual computations. Moreover, the type checking,
unwrapping and method lookups are often loop invariant and performance could be increased
by moving those operations out of the loop. We propose a simple scheme
to make a tracing JIT loop-aware by allowing it's existing optimizations to
perform loop invariant code motion.

@@

This is the key insight of the proposed implementation scheme: Giving an
optimization two iterations together at the same time gives the optimization
enough context to remove operations from the peeled loop, because it detects
that the operation was performed in the preamble already. Thus at runtime these
moved operations are only executed once when entering the loop and the results
are reused in further iterations.

@@

Before a trace is passed to the backend compiling it into machine code
it is optimized to achieve better performance.
One goal of that is to move 
operations out of the loop making them executed only once
and not every iteration. This we propose to achieve by loop peeling. It
leaves the loop body intact, but prefixes it with one iteration of the
loop. This operation by itself will not achieve anything. But if it is
combined with other optimizations it can increase the effectiveness of
those optimizations. For many optimization of interest only a few
additional details has to be considered when they are combined with loop peeling. These are
described below by explaining the loop peeling optimization
followed by a set of other optimizations and how they interact with
loop peeling.

@@

No special concerns needs to be taken when implementing redundant
guard removal together with loop peeling. The guards from
the preamble might make the guards of the peeled loop
redundant and thus removed. Therefore one effect of combining redundant
guard removal with loop peeling is that loop-invariant guards are moved out of the
loop. The peeled loop of the example reduces to

>>>>

\begin{abstract}
One of the nice properties of a tracing JIT is that many of its optimizations
are simple requiring one forward pass only. This is not true for loop-invariant code
motion which is a very important optimization for code with tight kernels.
Especially for dynamic languages that typically perform quite a lot of loop invariant
type checking, boxed value unwrapping and virtual method lookups.
In this paper we present a scheme for making simple optimizations loop-aware by
using a simple pre-processing step on the trace and not changing the
optimizations themselves. The scheme can give performance improvements of a
factor over two for PyPy's Python JIT executing simple numerical kernels
bringing the performance close to that of compiled C code.
\end{abstract}

@@

A dynamic language typically needs to do quite a lot of type
checking, wrapping/unwrapping of boxed values, and virtual method dispatching. 
For tight computationally intensive loops a
significant amount of the execution time might be spent on such tasks
instead of the actual computations. Moreover, the type checking,
unwrapping and method lookups are often loop invariant and performance could be increased
by moving those operations out of the loop. We propose a simple scheme
to make a tracing JIT loop-aware by allowing it's existing optimizations to
perform loop invariant code motion.

@@

This is the key insight of the proposed implementation scheme: If an
optimization is given two iterations together at the same time, the
optimization has enough context to remove operations from the peeled loop,
because it detects
that the operation was performed in the preamble already. Thus at runtime these
moved operations are only executed once when entering the loop and the results
are reused in further iterations.

@@

Before a trace is passed to the backend compiling it into machine code
it is optimized to achieve better performance.
One goal of that is to move 
operations out of the loop making them executed only once
and not every iteration. We propose to achieve this by loop peeling. It
leaves the loop body intact, but prefixes it with one iteration of the
loop. This operation by itself will not achieve anything. But if it is
combined with other optimizations it can increase the effectiveness of
those optimizations. For many optimizations of interest only a few
additional details have to be considered when they are combined with loop peeling. These are
described below by explaining the loop peeling optimization
followed by a set of other optimizations and how they interact with
loop peeling.

@@

No special concern needs to be taken when implementing redundant
guard removal together with loop peeling. The guards from
the preamble might make the guards of the peeled loop
redundant and thus removed. Therefore one effect of combining redundant
guard removal with loop peeling is that loop-invariant guards are moved out of the
loop. The peeled loop of the example reduces to
